Reviving stochasticity: uncertainty in SMBH binary eccentricity is unavoidable

### 2023-09-20 papers/paper-eccentricity/paper-eccentricity.tex
Final version of Reviving Stochasticity paper

%% Based on file 'sample631.tex'
%%
%% Modified 2021 March
%%
%% This is a sample manuscript marked up using the
%% AASTeX v6.31 LaTeX 2e macros.
%%
%% AASTeX is now based on Alexey Vikhlinin's emulateapj.cls 
%% (Copyright 2000-2015).  See the classfile for details.

%% AASTeX requires revtex4-1.cls and other external packages such as
%% latexsym, graphicx, amssymb, longtable, and epsf.  Note that as of 
%% Oct 2020, APS now uses revtex4.2e for its journals but remember that 
%% AASTeX v6+ still uses v4.1. All of these external packages should 
%% already be present in the modern TeX distributions but not always.
%% For example, revtex4.1 seems to be missing in the linux version of
%% TexLive 2020. One should be able to get all packages from www.ctan.org.
%% In particular, revtex v4.1 can be found at 
%% https://www.ctan.org/pkg/revtex4-1.

%% The first piece of markup in an AASTeX v6.x document is the \documentclass
%% command. LaTeX will ignore any data that comes before this command. The 
%% documentclass can take an optional argument to modify the output style.
%% The command below calls the preprint style which will produce a tightly 
%% typeset, one-column, single-spaced document.  It is the default and thus
%% does not need to be explicitly stated.
%%
%% using aastex version 6.3
\documentclass[twocolumn]{aastex631}

%% The default is a single spaced, 10 point font, single spaced article.
%% There are 5 other style options available via an optional argument. They
%% can be invoked like this:
%%
%% \documentclass[arguments]{aastex631}
%% 
%% where the layout options are:
%%
%%  twocolumn   : two text columns, 10 point font, single spaced article.
%%                This is the most compact and represent the final published
%%                derived PDF copy of the accepted manuscript from the publisher
%%  manuscript  : one text column, 12 point font, double spaced article.
%%  preprint    : one text column, 12 point font, single spaced article.  
%%  preprint2   : two text columns, 12 point font, single spaced article.
%%  modern      : a stylish, single text column, 12 point font, article with
%% 		  wider left and right margins. This uses the Daniel
%% 		  Foreman-Mackey and David Hogg design.
%%  RNAAS       : Supresses an abstract. Originally for RNAAS manuscripts 
%%                but now that abstracts are required this is obsolete for
%%                AAS Journals. Authors might need it for other reasons. DO NOT
%%                use \begin{abstract} and \end{abstract} with this style.
%%
%% Note that you can submit to the AAS Journals in any of these 6 styles.
%%
%% There are other optional arguments one can invoke to allow other stylistic
%% actions. The available options are:
%%
%%   astrosymb    : Loads Astrosymb font and define \astrocommands. 
%%   tighten      : Makes baselineskip slightly smaller, only works with 
%%                  the twocolumn substyle.
%%   times        : uses times font instead of the default
%%   linenumbers  : turn on lineno package.
%%   trackchanges : required to see the revision mark up and print its output
%%   longauthor   : Do not use the more compressed footnote style (default) for 
%%                  the author/collaboration/affiliations. Instead print all
%%                  affiliation information after each name. Creates a much 
%%                  longer author list but may be desirable for short 
%%                  author papers.
%% twocolappendix : make 2 column appendix.
%%   anonymous    : Do not show the authors, affiliations and acknowledgments 
%%                  for dual anonymous review.
%%
%% these can be used in any combination, e.g.
%%
%% \documentclass[twocolumn,linenumbers,trackchanges]{aastex631}
%%
%% AASTeX v6.* now includes \hyperref support. While we have built in specific
%% defaults into the classfile you can manually override them with the
%% \hypersetup command. For example,
%%
%% \hypersetup{linkcolor=red,citecolor=green,filecolor=cyan,urlcolor=magenta}
%%
%% will change the color of the internal links to red, the links to the
%% bibliography to green, the file links to cyan, and the external links to
%% magenta. Additional information on \hyperref options can be found here:
%% https://www.tug.org/applications/hyperref/manual.html#x1-40003
%%
%% Note that in v6.3 "bookmarks" has been changed to "true" in hyperref
%% to improve the accessibility of the compiled pdf file.
%%
%% If you want to create your own macros, you can do so
%% using \newcommand. Your macros should appear before
%% the \begin{document} command.
%%
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm} % italic bold math symbols

\newcommand{\vdag}{(v)^\dagger}
\newcommand\aastex{AAS\TeX}
\newcommand\latex{La\TeX}

\newcommand{\ketju}{\textsc{Ketju}}
\newcommand{\mstar}{\textsc{MSTAR}}
\newcommand{\gadget}{\textsc{Gadget-4}}

\newcommand{\Msun}{\ensuremath{\mathrm{M}_{\sun}}}
\newcommand{\Rvir}{\ensuremath{R_\mathrm{vir}}}

@@

% drafting macros
\newcommand{\drafting}[1]{
{\leavevmode\color[RGB]{224, 77, 24}#1}
}

\newcommand{\phj}[1]{
{\leavevmode\color[RGB]{0,0,255}#1}
}

%% Reintroduced the \received and \accepted commands from AASTeX v5.2
%\received{March 1, 2021}
%\revised{April 1, 2021}
%\accepted{\today}

%% Command to document which AAS Journal the manuscript was submitted to.
%% Adds "Submitted to " the argument.
%\submitjournal{PSJ}

%% For manuscript that include authors in collaborations, AASTeX v6.31
%% builds on the \collaboration command to allow greater freedom to 
%% keep the traditional author+affiliation information but only show
%% subsets. The \collaboration command now must appear AFTER the group
%% of authors in the collaboration and it takes TWO arguments. The last
%% is still the collaboration identifier. The text given in this
%% argument is what will be shown in the manuscript. The first argument
%% is the number of author above the \collaboration command to show with
%% the collaboration text. If there are authors that are not part of any
%% collaboration the \nocollaboration command is used. This command takes
%% one argument which is also the number of authors above to show. A
%% dashed line is shown to indicate no collaboration. This example manuscript
%% shows how these commands work to display specific set of authors 
%% on the front page.
%%
%% For manuscript without any need to use \collaboration the 
%% \AuthorCollaborationLimit command from v6.2 can still be used to 
%% show a subset of authors.
%
%\AuthorCollaborationLimit=2
%
%% will only show Schwarz & Muench on the front page of the manuscript
%% (assuming the \collaboration and \nocollaboration commands are
%% commented out).
%%
%% Note that all of the author will be shown in the published article.
%% This feature is meant to be used prior to acceptance to make the
%% front end of a long author article more manageable. Please do not use
%% this functionality for manuscripts with less than 20 authors. Conversely,
%% please do use this when the number of authors exceeds 40.
%%
%% Use \allauthors at the manuscript end to show the full author list.
%% This command should only be used with \AuthorCollaborationLimit is used.

%% The following command can be used to set the latex table counters.  It
%% is needed in this document because it uses a mix of latex tabular and
%% AASTeX deluxetables.  In general it should not be needed.
%\setcounter{table}{1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%
%% The following section outlines numerous optional output that
%% can be displayed in the front matter or as running meta-data.
%%
%% If you wish, you may supply running head information, although
%% this information may be modified by the editorial offices.
\shorttitle{Reviving Stochasticity}
\shortauthors{A. Rawlings et al.}
%%
%% You can add a light gray and diagonal water-mark to the first page 
%% with this command:
%% \watermark{text}
%% where "text", e.g. DRAFT, is the text to appear.  If the text is 
%% long you can control the water-mark size with:
%% \setwatermarkfontsize{dimension}
%% where dimension is any recognized LaTeX dimension, e.g. pt, in, etc.
%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\graphicspath{{./}{figures/}}
%% This is the end of the preamble.  Indicate the beginning of the
%% manuscript itself with \begin{document}.

\begin{document}

\title{Reviving stochasticity: uncertainty in SMBH binary eccentricity is unavoidable}

%% LaTeX will automatically break titles if they run longer than
%% one line. However, you may use \\ to force a line break if
%% you desire. In v6.31 you can include a footnote in the title.

%% A significant change from earlier AASTEX versions is in the structure for 
%% calling author and affiliations. The change was necessary to implement 
%% auto-indexing of affiliations which prior was a manual process that could 
%% easily be tedious in large author manuscripts.
%%
%% The \author command is the same as before except it now takes an optional
%% argument which is the 16 digit ORCID. The syntax is:
%% \author[xxxx-xxxx-xxxx-xxxx]{Author Name}
%%
%% This will hyperlink the author name to the author's ORCID page. Note that
%% during compilation, LaTeX will do some limited checking of the format of
%% the ID to make sure it is valid. If the "orcid-ID.png" image file is 
%% present or in the LaTeX pathway, the OrcID icon will appear next to
%% the authors name.
%%
%% Use \affiliation for affiliation information. The old \affil is now aliased
%% to \affiliation. AASTeX v6.31 will automatically index these in the header.
%% When a duplicate is found its index will be the same as its previous entry.
%%
%% Note that \altaffilmark and \altaffiltext have been removed and thus 
%% can not be used to document secondary affiliations. If they are used latex
%% will issue a specific error message and quit. Please use multiple 
%% \affiliation calls for to document more than one affiliation.
%%
%% The new \altaffiliation can be used to indicate some secondary information
%% such as fellowships. This command produces a non-numeric footnote that is
%% set away from the numeric \affiliation footnotes.  NOTE that if an
%% \altaffiliation command is used it must come BEFORE the \affiliation call,
%% right after the \author command, in order to place the footnotes in
%% the proper location.
%%
%% Use \email to set provide email addresses. Each \email will appear on its
%% own line so you can put multiple email address in one \email call. A new
%% \correspondingauthor command is available in V6.31 to identify the
%% corresponding author of the manuscript. It is the author's responsibility
%% to make sure this name is also in the author list.
%%
%% While authors can be grouped inside the same \author and \affiliation
%% commands it is better to have a single author for each. This allows for
%% one to exploit all the new benefits and should make book-keeping easier.
%%
%% If done correctly the peer review system will be able to
%% automatically put the author and affiliation information from the manuscript
%% and save the corresponding author the trouble of entering it by hand.

%\correspondingauthor{August Muench}
%\email{greg.schwarz@aas.org, gus.muench@aas.org}

\author[0000-0003-1807-6321]{Alexander Rawlings}
\affiliation{Department of Physics,
Gustaf H\"allstr\"omin katu 2, FI-00014, University of Helsinki, Finland}

\author[0000-0001-5721-9335]{Matias Mannerkoski}
\affiliation{Department of Physics,
Gustaf H\"allstr\"omin katu 2, FI-00014, University of Helsinki, Finland}

\author[0000-0001-8741-8263]{Peter H. Johansson}
\affiliation{Department of Physics,
Gustaf H\"allstr\"omin katu 2, FI-00014, University of Helsinki, Finland}

\author[0000-0002-7314-2558]{Thorsten Naab}
\affiliation{Max-Planck-Institut f\"ur Astrophysik, Karl-Schwarzchild-Str 1,
D-85748 Garching, Germany}

\correspondingauthor{Alexander Rawlings}
\email{alexander.rawlings@helsinki.fi}

%% Note that the \and command from previous versions of AASTeX is now
%% depreciated in this version as it is no longer necessary. AASTeX 
%% automatically takes care of all commas and "and"s between authors names.

%% AASTeX 6.31 has the new \collaboration and \nocollaboration commands to
%% provide the collaboration status of a group of authors. These commands 
%% can be used either before or after the list of corresponding authors. The
%% argument for \collaboration is the collaboration identifier. Authors are
%% encouraged to surround collaboration identifiers with ()s. The 
%% \nocollaboration command takes no argument and exists to indicate that
%% the nearby authors are not part of surrounding collaborations.

%% Mark off the abstract in the ``abstract'' environment. 
\begin{abstract}
We study supermassive black hole (SMBH) binary eccentricity of equal-mass galaxy mergers in $N$-body simulations with the \ketju{} code, which combines the \gadget{} fast multipole gravity solver with accurate regularized integration and post-Newtonian corrections around SMBHs.
In simulations with realistic, high eccentricity galactic merger orbits, the binary eccentricity is found to be a non-linear function of the deflection angle in the SMBH orbit during the final, nearly radial close encounter between the SMBHs before they form a bound binary.
This mapping between the deflection angle and the binary eccentricity has no apparent resolution dependence in our simulations spanning the resolution range of $1\times10^5$--$8\times10^6$ particles per galaxy.
The mapping is also captured using a simple model with an analytic potential, indicating that it is driven by the interplay between a smooth asymmetric stellar background potential and dynamical friction acting on the SMBHs.
Due to the non-linearity of this mapping, in certain merger configurations small, parsec-scale variations in the merger orbit can result in binary eccentricities varying in nearly the full possible range between $e=0$ and $e=1$.
In idealized simulations, such variations are caused by finite resolution effects,
and convergence of the binary eccentricity can be achieved with increasing resolution.
However, in real galaxies, other mechanisms such as nuclear gas and substructure that perturb the merger orbit are likely to be significant enough for the binary eccentricity to be effectively random.
Our results indicate that the distribution of these effectively random eccentricities can be studied using even moderate resolution simulations.

\end{abstract}

%% Keywords should appear after the \end{abstract} command. 
%% The AAS Journals now uses Unified Astronomy Thesaurus concepts:
%% https://astrothesaurus.org
%% You will be asked to selected these concepts during the submission process
%% but this old "keyword" functionality is maintained in case authors want
%% to include these concepts in their preprints.
%\keywords{Galaxy mergers (608); Supermassive black holes (1663); Eccentricity (441); N-body simulations (1083); }

%% From the front matter, we move on to the body of the paper.
%% Sections are demarcated by \section and \subsection, respectively.
%% Observe the use of the LaTeX \label
%% command after the \subsection to give a symbolic KEY to the
%% subsection for cross-referencing in a \ref command.
%% You can use LaTeX's \ref and \label commands to keep track of
%% cross-references to sections, equations, tables, and figures.
%% That way, if you change the order of any elements, LaTeX will
%% automatically renumber them.
%%
%% We recommend that authors also use the natbib \citep
%% and \citet commands to identify citations.  The citations are
%% tied to the reference list via symbolic KEYs. The KEY corresponds
%% to the KEY in the \bibitem in the reference list below.

%-------------------------- MAIN TEXT STARTS HERE --------------------------%
% INTRODUCTION
\section{Introduction}
Supermassive black holes (SMBHs) are believed to reside at the centers of all massive galaxies \citep[e.g.][]{kormendy2013}.
In the $\Lambda$CDM model as galaxies grow through gas accretion and mergers \citep[e.g.][]{volonteri2003,naab2017} their SMBHs 
are also expected to interact in a three-phase merger process \citep{begelman1980}.

Firstly, dynamical friction \citep[DF,][]{chandrasekhar1943} acts to bring the SMBHs from kiloparsec-scales down to parsec-scale separations, after which the SMBHs form a bound binary with a semimajor axis $a$ and eccentricity $e$ \citep[e.g.][]{milosavljevic2001,merritt2013}. In the second phase, the
SMBH binary separation is reduced through sequential slingshot encounters with the surrounding stellar
distribution \citep{hills1980, hills1983, quinlan1996,rantala2018}. Finally, at small subparsec separations gravitational wave (GW) emission
becomes the dominant mechanism by which the SMBH binary can lose its remaining orbital energy and angular momentum, thus driving the
SMBHs to coalescence, where the strength of the GW emission strongly depends on the
binary eccentricity \citep{peters1963,peters1964,mannerkoski2019}.

The complex nature of SMBH coalescence in a galaxy merger setting necessitates the use of numerical techniques \citep[e.g.][]{berentzen2009,khan2011,dosopoulou2017,mannerkoski2019}
to provide quantitative predictions for ongoing observational programs such as ground-based pulsar timing arrays \citep[PTAs,][]{agazie2023,2023Antoniadis,2023Xu,zic2023}
and the upcoming Laser Interferometer Space Antenna \citep[LISA, e.g.][]{amaro-seoane2023}.
Understanding how faithfully SMBH binary eccentricity is captured in numerical simulations is thus critical in constraining the SMBH merger rate
that the observational GW community aims to detect.

The SMBH binary merging process and its dependence on eccentricity and resolution has been extensively studied using collisionsless merger simulations \citep[e.g.][]{berentzen2009,vasiliev2015,bortolas2016,gualandris2017,gualandris2022}. 
Recently, \citet{nasim2020} have argued that the substantial scatter in SMBH binary eccentricity observed in gas-free merger simulations is an artefact of poor phase space
discretization, and that in the real Universe where SMBH masses are far greater than stellar masses $(M_\bullet \gg m_\star)$, SMBH binary eccentricity is a reliably predictable quantity.

In this Letter, we find that for realistic galaxy merger orbits the scatter in the final SMBH binary eccentricity is due to the 
physical sensitivity of the final eccentricity to small perturbations of the final  nearly radial plunging trajectory of the SMBHs before they become bound, even in the infinite resolution limit.
As a result, we argue that the stochasticity of the binary eccentricity is an unavoidable physical feature of realistic galaxy mergers, at least in the equal-mass case.

\begin{figure*}
    \centering
    \includegraphics{eccentricities.pdf}
    \caption{Binary eccentricity $e$ (note the non-linear scale) and shifted time $t'=t-t_\mathrm{bound}$ for the $M_\bullet/m_\star=5000$ resolution simulations from the $e_0 = 0.90$ set (blue lines) and $e_0 = 0.99$ set (orange lines). 
    The downwards-pointing arrows indicate the median time when the SMBH binaries becomes hard. 
    In the $e_0=0.90$ set, six simulations have a binary merger within $50\,\mathrm{Myr}$.}
    \label{fig:eccentricities}
\end{figure*}

% NUMERICAL
\section{Numerical Simulations}

We construct a number of idealized galaxy merger simulations, which we evolve with our new version of the \ketju{} code \citep{mannerkoski2023,rantala2017}.
The dynamics of
SMBHs, and stars in a small region around them, are integrated with an algorithmically regularized integrator \citep{rantala2020}, whereas the dynamics of the remaining particles
is computed with the \gadget{} \citep{springel2021} fast multiple method (FMM) with second order multipoles. Together with hierarchical time integration this allows for
symmetric interactions and manifest momentum conservation. \ketju{} also includes post-Newtonian (PN) correction terms up to order 3.5 between each pair of SMBHs \citep{blanchet2014}.

Our galaxy models represent the nuclear bulge of a gas-devoid elliptical galaxy, matching the models of \citet{nasim2020}.
Each galaxy consists of a stellar-only \citet{dehnen1993} sphere with shape parameter $\gamma=0.5$ and scale radius $a=186\,\mathrm{pc}$, where the Dehnen profile is given by:
\begin{equation}\label{eq:dehnen}
    \rho_\star(r) = \frac{(3-\gamma)M_\star}{4\pi} \frac{a}{r^\gamma (r+a)^{(4-\gamma)}}.
\end{equation}
The total stellar mass is $M_\star=10^{10}\,\Msun$, and at the center of the model galaxy a SMBH of mass $M_\bullet=10^8\,\Msun$ is placed.
We test seven different mass resolutions 
with a varying number of stellar particles: $N_\star = \{1.0\times10^5,2.5\times10^5, 5.0\times10^5, 1.0\times10^6, 2.0\times10^6, 4.0\times10^6, 8.0\times10^6\}$, corresponding to $M_\bullet/m_\star= 1000 \textnormal{--} 80000$.

We then construct isolated merger initial conditions by placing two galaxies on two different elliptical orbits with eccentricities of $e_0=0.90$ and $e_0=0.99$, at a fixed
initial separation of $D=3.72\,\mathrm{kpc}$ and a semimajor axis of $a_0=2.79\,\mathrm{kpc}$.
The $e_0=0.90$ merger orbit matches that used by \citet{nasim2020}, with the radial and tangential velocities being consistent with the values reported in \citet{gualandris2022}.
For each orbital configuration, we run ten realizations for each mass resolution, to account for stochasticity caused by the discretized phase space. 
Interactions between stellar particles are softened with a softening length of $\varepsilon = 2.5\,\mathrm{pc}$, and the \ketju{} region radius is $r_{\rm ketju}=3\varepsilon = 7.5\,\mathrm{pc}$.

@@

\begin{figure}
    \includegraphics{convergence}
    \caption{Convergence of the eccentricity scatter $\sigma_e$ as a function of mass resolution $M_\bullet/m_\star$ and the total number of stellar particles $N_{\star, \mathrm{tot}}$. 
    The expected $1/\sqrt{N_\star}$ scaling is recovered for the $e_0 = 0.90$ mergers. 
    For the $e_0=0.99$ mergers, $\sigma_e$ does not decrease until $M_\bullet/m_\star \geq 20000$.}
    \label{fig:convergence}
\end{figure}

@@

For the $e_0=0.90$ orbit, the convergence of $\sigma_e$ scales as $1/\sqrt{N_{\star,\mathrm{tot}}}$, where $N_{\star,\mathrm{tot}}$ is the total number of stellar particles in the merger.
The values of $\sigma_e$ we report are quantitatively similar to the values found by \citet{nasim2020} for the same system. 
The variation in eccentricity does not show the same scaling for the $e_0=0.99$ orbit as the $e_0=0.90$ orbit. 
For mass resolutions $M_\bullet/m_\star \leq 10000$, the value of $\sigma_e$ is almost constant, before significantly dropping at higher mass resolutions.

>>>>

% mnras_template.tex 
%
% LaTeX template for creating an MNRAS paper
%
% v3.0 released 14 May 2015
% (version numbers match those of mnras.cls)
%
% Copyright (C) Royal Astronomical Society 2015
% Authors:
% Keith T. Smith (Royal Astronomical Society)

% Change log
%
% v3.0 May 2015
%    Renamed to match the new package name
%    Version number matches mnras.cls
%    A few minor tweaks to wording
% v1.0 September 2013
%    Beta testing only - never publicly released
%    First version: a simple (ish) template for creating an MNRAS paper

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Basic setup. Most papers should leave these options alone.
\documentclass[fleqn,usenatbib]{mnras}

% MNRAS is set in Times font. If you don't have this installed (most LaTeX
% installations will be fine) or prefer the old Computer Modern fonts, comment
% out the following line
\usepackage{newtxtext,newtxmath}
% Depending on your LaTeX fonts installation, you might get better results with one of these:
%\usepackage{mathptmx}
%\usepackage{txfonts}

% Use vector fonts, so it zooms properly in on-screen viewing software
% Don't change these lines unless you know what you are doing
\usepackage[T1]{fontenc}

% Allow "Thomas van Noord" and "Simon de Laguarde" and alike to be sorted by "N" and "L" etc. in the bibliography.
% Write the name in the bibliography as "\VAN{Noord}{Van}{van} Noord, Thomas"
\DeclareRobustCommand{\VAN}[3]{#2}
\let\VANthebibliography\thebibliography
\def\thebibliography{\DeclareRobustCommand{\VAN}[3]{##3}\VANthebibliography}

%%%%% AUTHORS - PLACE YOUR OWN PACKAGES HERE %%%%%

% Only include extra packages if you really need them. Common packages are:
\usepackage{graphicx}	% Including figure files
\usepackage{amsmath}	% Advanced maths commands
%\usepackage{amssymb}	% Extra maths symbols
\usepackage{bm}         % Bold math fonts

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%% AUTHORS - PLACE YOUR OWN COMMANDS HERE %%%%%

% Please keep new commands to a minimum, and use \newcommand not \def to avoid
% overwriting existing commands. Example:
%\newcommand{\pcm}{\,cm$^{-2}$}	% per cm-squared
\newcommand{\ketju}{\textsc{Ketju}}
\newcommand{\mstar}{\textsc{mstar}}
\newcommand{\gadget}{\textsc{gadget-4}}

\newcommand{\Msun}{\ensuremath{M_{\sun}}}

@@

\graphicspath{{figures/}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%% TITLE PAGE %%%%%%%%%%%%%%%%%%%

% Title of the paper, and the short title which is used in the headers.
% Keep the title short and informative.
\title[Reviving Stochasticity]{Reviving stochasticity: uncertainty in SMBH binary eccentricity is unavoidable}

% The list of authors, and the short list which is used in the headers.
% If you need two or more lines of authors, add an extra line using \newauthor
\author[A. Rawlings et al.]{
Alexander Rawlings,$^{1}$\thanks{E-mail: alexander.rawlings@helsinki.fi}
Matias Mannerkoksi,$^{1}$
Peter H. Johansson,$^{1}$
Thorsten Naab$^{2}$
\vspace*{0.1cm}\\%
% List of institutions
$^{1}$
Department of Physics,
Gustaf H\"allstr\"omin katu 2, FI-00014, University of Helsinki, Finland
\\%
$^{2}$
Max-Planck-Institut f\"ur Astrophysik, Karl-Schwarzchild-Str 1, D-85748 Garching, Germany
}

% These dates will be filled out by the publisher
\date{Accepted XXX. Received YYY; in original form ZZZ}

% Enter the current year, for the copyright statements etc.
\pubyear{2023}

% Don't change these lines
\begin{document}
\label{firstpage}
\pagerange{\pageref{firstpage}--\pageref{lastpage}}
\maketitle

% Abstract of the paper
\begin{abstract}
We study supermassive black hole (SMBH) binary eccentricity of equal-mass galaxy mergers in $N$-body simulations with the \ketju{} code, which combines the \gadget{} fast multipole gravity solver with accurate regularised integration and post-Newtonian corrections around SMBHs.
In simulations with realistic, high eccentricity galactic merger orbits, the hard binary eccentricity is found to be a non-linear function of the deflection angle in the SMBH orbit during the final, nearly radial close encounter between the SMBHs before they form a bound binary.
This mapping between the deflection angle and the binary eccentricity has no apparent resolution dependence in our simulations spanning the resolution range of $1\times10^5$--$8\times10^6$ particles per galaxy.
The mapping is also captured using a simple model with an analytic potential, indicating that it is driven by the interplay between a smooth asymmetric stellar background potential and dynamical friction acting on the SMBHs.
Due to the non-linearity of this mapping, in eccentric major merger configurations small, parsec-scale variations in the merger orbit can result in binary eccentricities varying in nearly the full possible range between $e=0$ and $e=1$.
In idealised simulations, such variations are caused by finite resolution effects,
and convergence of the binary eccentricity can be achieved with increasing resolution.
However, in real galaxies, other mechanisms such as nuclear gas and substructure that perturb the merger orbit are likely to be significant enough for the binary eccentricity to be effectively random.
Our results indicate that the distribution of these effectively random eccentricities can be studied using even moderate resolution simulations.
\end{abstract}

% Select between one and six entries from the list of approved keywords.
% Don't make up new ones.
\begin{keywords}
black hole physics -- galaxies: kinematics and dynamics -- methods: numerical -- software: simulations
\end{keywords}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%% BODY OF PAPER %%%%%%%%%%%%%%%%%%

% INTRODUCTION
\section{Introduction}
Supermassive black holes (SMBHs) are believed to reside at the centres of all massive galaxies \citep[e.g.][]{kormendy2013}.
In the $\Lambda$CDM model as galaxies grow through gas accretion and mergers \citep[e.g.][]{volonteri2003,naab2017} their SMBHs 
are also expected to interact in a three-phase merger process \citep{begelman1980}.

Firstly, dynamical friction \citep[DF,][]{chandrasekhar1943} acts to bring the SMBHs from kiloparsec-scales down to parsec-scale separations, after which the SMBHs form a bound binary with a semimajor axis $a$ and eccentricity $e$ \citep[e.g.][]{milosavljevic2001,merritt2013}. In the second phase, the
SMBH binary separation is reduced through sequential slingshot encounters with the surrounding stellar
distribution \citep{hills1980, hills1983, quinlan1996,rantala2018}. Finally, at small subparsec separations gravitational wave (GW) emission
becomes the dominant mechanism by which the SMBH binary can lose its remaining orbital energy and angular momentum, thus driving the
SMBHs to coalescence \citep{peters1963,peters1964}.

The complex nature of SMBH coalescence in a galaxy merger setting necessitates the use of numerical techniques \citep[e.g.][]{berentzen2009,khan2011,dosopoulou2017,mannerkoski2019}
to provide quantitative predictions for ongoing observational programmes such as ground-based pulsar timing arrays \citep[PTAs,][]{agazie2023,2023Antoniadis,2023Xu,zic2023}
and the upcoming Laser Interferometer Space Antenna \citep[LISA, e.g.][]{amaro-seoane2023}.
In particular, the eccentricity of the binary significantly affects the GW emission, with higher eccentricities resulting both in faster mergers as well as changes in the emitted GW spectrum \citep{enoki2007, huerta2015, taylor2016}.
Understanding SMBH binary eccentricity and how faithfully it is captured in numerical simulations is thus critical for predicting and interpreting observations done with instruments such as PTAs and LISA.

The SMBH binary merging process and its dependence on eccentricity and resolution has been extensively studied using collisionless merger simulations \citep[e.g.][]{berentzen2009,vasiliev2015,bortolas2016,gualandris2017,gualandris2022}. 
Recently, \citet{nasim2020} have argued that the substantial scatter in SMBH binary eccentricity observed in gas-free merger simulations is an artefact of poor phase space
discretisation, and that in the real Universe where SMBH masses are far greater than stellar masses $(M_\bullet \gg m_\star)$, SMBH binary eccentricity is a reliably predictable quantity.

In this paper, we find that for realistic galaxy merger orbits the scatter in the final SMBH binary eccentricity is due to the 
physical sensitivity of the final eccentricity to small perturbations of the final  nearly radial plunging trajectory of the SMBHs before they become bound, even in the infinite resolution limit.
As a result, we argue that the stochasticity of the binary eccentricity is an unavoidable physical feature of realistic galaxy mergers, at least in the equal-mass case.

This paper is structured as follows. In Section \ref{sec:num_sims} we briefly discuss the main features
of the \ketju{} code and our numerical simulations. In Section \ref{sec:results} we study the eccentricity scatter in our simulations and also present a simple model that captures the observed scatter. 
We discuss our results and their implications in Section \ref{sec:discussion} and finally present our conclusions in Section \ref{sec:conclusions}.

\begin{figure*}
    \centering
    \includegraphics{eccentricities.pdf}
    \caption{Binary eccentricity $e$ (note the non-linear scale) and shifted time $t'=t-t_\mathrm{bound}$ for the $M_\bullet/m_\star=5000$ resolution simulations from the $e_0 = 0.90$ set (blue lines) and $e_0 = 0.99$ set (orange lines). 
    The circle markers indicate the time when a hard binary formed, with the preceding semitransparent lines indicating the transition period when the hard binary is still forming, and the solid lines the hard binary evolution. 
    During the transition period, the Keplerian definition of binary motion is affected by the intervening stellar mass, creating the oscillatory artefacts in the semitransparent lines.
    In the $e_0=0.90$ simulation set, six simulations have a binary merger within $50\,\mathrm{Myr}$.}
    \label{fig:eccentricities}
\end{figure*}

% NUMERICAL
\section{Numerical Simulations}
\label{sec:num_sims}

We construct a number of idealised galaxy merger simulations, which we evolve with our new version of the \ketju{} code \citep{mannerkoski2023,rantala2017}.
The dynamics of
SMBHs, and stars in a small region around them, are integrated with an algorithmically regularised integrator \citep{rantala2020}, whereas the dynamics of the remaining particles
is computed with the \gadget{} \citep{springel2021} fast multiple method (FMM) with second order multipoles. Together with hierarchical time integration this allows for
symmetric interactions and manifest momentum conservation. \ketju{} also includes post-Newtonian (PN) correction terms up to order 3.5 between each pair of SMBHs \citep{blanchet2014}.

Our galaxy models represent the nuclear bulge of a gas-devoid elliptical galaxy, and are designed to match exactly the models of \citet{nasim2020}.
Each galaxy consists of a stellar-only \citet{dehnen1993} sphere with shape parameter $\gamma=0.5$ and scale radius $a=186\,\mathrm{pc}$, where the Dehnen profile is given by:
\begin{equation}\label{eq:dehnen}
    \rho_\star(r) = \frac{(3-\gamma)M_\star}{4\pi} \frac{a}{r^\gamma (r+a)^{(4-\gamma)}}.
\end{equation}
The total stellar mass is $M_\star=10^{10}\,\Msun$, and at the centre of the model galaxy a SMBH of mass $M_\bullet=10^8\,\Msun$ is placed\footnote{We set the length and mass scales of the system to match one of those presented in \citet{gualandris2022}. However, as the performed simulations are gravity-only, the system may be transformed to other mass and length scales with the same $M_\mathrm{tot}/a$ ratio without affecting the results.}.
Our galaxies lie on the observed $M_\bullet\text{--}\sigma$ relation presented in \citet{vandenbosch2016}.
We test seven different mass resolutions 
with a varying number of stellar particles: $N_\star = \{1.0\times10^5,2.5\times10^5, 5.0\times10^5, 1.0\times10^6, 2.0\times10^6, 4.0\times10^6, 8.0\times10^6\}$, corresponding to $M_\bullet/m_\star= 1000 \textnormal{--} 80000$.

We then construct isolated merger initial conditions by placing two galaxies on two different elliptical orbits with eccentricities of $e_0=0.90$ and $e_0=0.99$, at a fixed
initial separation of $D=3.72\,\mathrm{kpc}$ and a semimajor axis of $a_0=2.79\,\mathrm{kpc}$.
The $e_0=0.90$ merger orbit matches that used by \citet{nasim2020}, with the radial and tangential velocities being consistent with the values reported in \citet{gualandris2022}.
For each orbital configuration, we run ten realisations for each mass resolution, to account for stochasticity caused by the discretised phase space. 
Interactions between stellar particles are softened with a softening length of $\varepsilon = 2.5\,\mathrm{pc}$, and the \ketju{} region radius is set to $r_{\rm ketju}=3\varepsilon = 7.5\,\mathrm{pc}$.

@@

\begin{figure}
    \includegraphics{convergence}
    \caption{Convergence of the eccentricity scatter $\sigma_e$ as a function of mass resolution $M_\bullet/m_\star$ and the total number of stellar particles $N_{\star, \mathrm{tot}}$. 
    The results from \citet{nasim2020} are shown as teal triangles for comparison.
    The expected $1/\sqrt{N_\star}$ scaling is recovered for the $e_0 = 0.90$ mergers. 
    For the $e_0=0.99$ mergers, $\sigma_e$ does not decrease until $M_\bullet/m_\star \geq 20000$.}
    \label{fig:convergence}
\end{figure}

@@

For the $e_0=0.90$ orbit, the convergence of $\sigma_e$ scales as $1/\sqrt{N_{\star,\mathrm{tot}}}$, where $N_{\star,\mathrm{tot}}$ is the total number of stellar particles in the merger.
The values of $\sigma_e$ we report are quantitatively similar to the values found by \citet{nasim2020} for the same system, as can be seen by the teal triangles in \autoref{fig:convergence}.
The variation in eccentricity does not show the same scaling for the $e_0=0.99$ orbit as the $e_0=0.90$ orbit. 
For mass resolutions $M_\bullet/m_\star \leq 10000$, the value of $\sigma_e$ is almost constant, before significantly dropping at higher mass resolutions.
